Hydra Battles and Cie

### 2021-09-22 doc/JFLA2022/paper.tex
small corrections
comparison with the earlier (2018) paper

\usepackage{doc}
\usepackage{tikz}
\usepackage{tikzsymbols}
\usetikzlibrary{arrows}

@@

\institute{
 LaBRI, Université Bordeaux \\
  \email{pierre.casteran@labri.fr}
\and
   Ask for agreement !
 }

@@

Hydra games (also known as \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians, 
Laurie Kirby and Jeff Paris: \emph{Accessible Independence Results for Peano Arithmetic}~\cite{KP82}.
This article describes a game between two players: Hercules and a hydra.
A short description  can be found in Andrej Bauer's blogpost~\cite{bauer2008} as well as a simulator~\cite{BauerHydra}.
In a few words:
\begin{itemize}
\item A hydra is a finite tree, traditionnaly presented with the root at the bottom, the leaves of which are called \emph{heads}
  (Fig~\vref{fig:iota-example}).
\item At every round, Hercules chops off one head of the hydra. If the head is at a distance greater or equal than 2 from the root,
  then some sub-tree $h$ of the hydra is copied a certain amount $i$ of times. The number $i$ of copies and the copied sub-hydra $h$ may depend of the considered variant of the game
  or the time elapsed since the beginning of the fight.
\end{itemize}
   \begin{figure}[htb]
\centering
\begin{tikzpicture}[very thick, scale=0.3]
\node (foot) at (10,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (10,2) {$\bullet$};
\node (N22) at (7,2) {$\bullet$};
\node (N3) at (14,2) {$\bullet$};
\node (N4) at (18,2) {$\Smiley[2][green]$};
\node (N5) at (0,4) {$\bullet$};
\node (N6) at (2,5) {$\Smiley[2][green]$};
\node (N7) at (4,6) {$\Smiley[2][green]$};
\node (N88) at (7,4) {$\bullet$};
\node (N8) at (10,4) {$\bullet$};
\node (N9) at (14,6) {$\Smiley[2][green]$};
\node (N10) at (0,8) {$\Smiley[2][green]$};
\node (N11) at (10,7) {$\Smiley[2][green]$};
\node (N111) at (7,7) {$\Smiley[2][green]$};
\draw (foot) to [bend left=10] (N1);
\draw (foot) -- (N2);
\draw (foot) -- (N22);
\draw (foot) -- (N3);
\draw (foot) -- (N4);
\draw (N1) to  (N5);
\draw (N1) to   [bend left=10] (N6);
\draw (N1) to   [bend right=20] (N7);
\draw (N2) to  (N8);
\draw (N22) to  (N88);
\draw (N8) to  (N11);
\draw (N88) to  (N111);
\draw (N3) to  (N9);
\draw (N5) to  (N10);
\end{tikzpicture}
\caption{The hydra associated with $\omega^{\omega+2}+\omega^\omega \times 2 + \omega + 1$ \label{fig:iota-example}}

@@

\begin{quote}
\begin{description}
\item[$(1)$] In the Hydra game, every strategy is a winning strategy for Hercules.
  \item[$(2)$] The theorem $(1)$ cannot be proved in Peano Arithmetic.
\end{description}
\end{quote}

@@

\begin{itemize}
\item We present complete proof scripts for the \textit{Coq} proof assistant~\cite{Coq}. These scripts may consulted and replayed by the reader.
\item Complex formalisations and proofs are explained in an
  electronic book~\cite{HydraBook}. Whenever various reasonable choices exist, we try to present and compare the alternatives.
  For instance, Figures~\vref{fig:Ex42E0} and \vref{fig:Ex42-schutte} show two radically different proofs of the equality
  $\omega+42+\omega^2=\omega^2$, the first one is a simple proof by computation, the second one shows how this equality
  is a consequence of the axioms of the set-theoretic model  by Kurt Schütte~\cite{schutte}. 
\item
  Reciprocally, we provide ``concrete'' examples of several formalisation and proof techniques: operational type classes, functions defined by  equations, dependently typed functions, etc.
\item It may be used as a library on ordinal numbers, for instance for proving termination properties.
  \end{itemize}

@@

\section{New results}
In order to make evident how long a hydra battle can be,
one may consider a deterministic variant of the game,
``compute'' the number of steps of the battles, and thus obtain a minoration of the general case.

The variant we consider is the same as 
in~\cite{KP82, bauer2008, BauerHydra}.

 \begin{itemize}
 \item The game starts at an arbitrary initial step $i\in\mathbb{N}$.

   \item The hydra grows $n$ copies at step $n$ of the game.
   \item  The initial hydra is the graphical representation of an ordinal  $\alpha<\epsilon_0$  in Cantor normal form (See for instance Fig~\ref{fig:iota-example}). This property is an invariant of the game.
 \item Hercules always chops off the rightmost head of the hydra (which is always one of the lowest heads).
 \end{itemize}
 It mathematical terms, if at step $n$ the hydra is associated with the ordinal $\alpha$, at step $n+1$ it is associated with
 $\canonseq{\alpha}{n+1}$, the $(n+1)$th element of the canonical sequence of $\alpha$~\cite{KS81}.

@@

  \item Presentation of the main steps of proofs, possibly interleaved with text (Figures \vref{fig:Ex42E0} and \vref{fig:Ex42-schutte}).

@@

Another direction would be to write in \textit{Coq} a formal proof of the original statement of the second theorem of Kirby and Paris. For this purpose, we plan to import  the formalization of Peano Arithmetic by O'Connor~\cite{Goedel}.

>>>>

\usepackage{doc}
\usepackage{tikz}
\usepackage{tikzsymbols}
\newtheorem{theorem}{Theorem}
\usetikzlibrary{arrows}

@@

\institute{
Univ. Bordeaux, CNRS, Bordeaux INP, LaBRI, UMR 5800, F-33400 Talence, France \\
  \email{pierre.casteran@labri.fr}
\and
   Ask for agreement !
 }

@@

Hydra games (also known as \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians, 
Laurie Kirby and Jeff Paris: \emph{Accessible Independence Results for Peano Arithmetic}~\cite{KP82}.
This article describes a game between two players: Hercules and a hydra.
A short description  can be found in Andrej Bauer's blogpost~\cite{bauer2008} as well as a simulator~\cite{BauerHydra}.
In a few words:
\begin{itemize}
\item A hydra is a finite tree, traditionnaly presented with the root at the bottom, the leaves of which are called \emph{heads}
  (Fig~\ref{fig:iota-example}).
\item At every round, Hercules chops off one head of the hydra. If the head is at a distance greater or equal than 2 from the root,
  then some sub-tree $h$ of the hydra is copied a certain amount $n$ of times. The number $n$ of copies and the copied sub-hydra $h$ may depend of the considered variant of the game
  or the time elapsed since the beginning of the fight.
\end{itemize}
   \begin{figure}[htb]
\centering
\begin{tikzpicture}[very thick, scale=0.3]
\node (foot) at (10,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (10,2) {$\bullet$};
\node (N22) at (7,2) {$\bullet$};
\node (N3) at (14,2) {$\bullet$};
\node (N4) at (18,2) {$\Smiley[2][green]$};
\node (N5) at (0,4) {$\bullet$};
\node (N6) at (2,5) {$\Smiley[2][green]$};
\node (N7) at (4,6) {$\Smiley[2][green]$};
\node (N88) at (7,4) {$\bullet$};
\node (N8) at (10,4) {$\bullet$};
\node (N9) at (14,6) {$\Smiley[2][green]$};
\node (N10) at (0,8) {$\Smiley[2][green]$};
\node (N11) at (10,7) {$\Smiley[2][green]$};
\node (N111) at (7,7) {$\Smiley[2][green]$};
\draw (foot) to [bend left=10] (N1);
\draw (foot) -- (N2);
\draw (foot) -- (N22);
\draw (foot) -- (N3);
\draw (foot) -- (N4);
\draw (N1) to  (N5);
\draw (N1) to   [bend left=10] (N6);
\draw (N1) to   [bend right=20] (N7);
\draw (N2) to  (N8);
\draw (N22) to  (N88);
\draw (N8) to  (N11);
\draw (N88) to  (N111);
\draw (N3) to  (N9);
\draw (N5) to  (N10);
\end{tikzpicture}
\caption{The hydra associated with $\omega^{\omega+2}+\omega^\omega \times 2 + \omega + 1$ \label{fig:iota-example}}

@@

\begin{theorem}
  In the Hydra game, every strategy is a winning strategy for Hercules. \label{kp:thm1}
\end{theorem}

\begin{theorem}
  Theorem~\ref{kp:thm1} cannot be proved in Peano Arithmetic. \label{kp:thm2}
\end{theorem}

@@

We present complete proof scripts for the \textit{Coq} proof assistant~\cite{Coq}. These scripts may consulted and replayed by the reader.

Complex formalisations and proofs are explained in an
  electronic book~\cite{HydraBook}. Whenever various reasonable choices exist, we try to present and compare the alternatives.
  For instance, Figures~\ref{fig:Ex42E0} and \ref{fig:Ex42-schutte} show two radically different proofs of the equality
  $\omega+42+\omega^2=\omega^2$, the first one is a simple proof by computation, the second one shows how this equality
  is a consequence of the axioms of the set-theoretic model  by Kurt Schütte~\cite{schutte}.

  Reciprocally, we provide ``concrete'' examples of several formalisation and proof techniques: operational type classes, functions defined by  equations, dependently typed functions, etc.

Our development may be used as a library on ordinal numbers, for instance for proving termination properties.

@@

\section{Changes in the library}
The 2018 article contains the proof of  a variant of Theorem~\ref{kp:thm2}:

\begin{theorem}
  There is no termination measure mapping hydras to a segment $[0,\mu)$ where $\mu<\epsilon_0$ for proving the termination of  all hydra battles.\label{thm3}
\end{theorem}

Indeed, the consideration of \emph{all} battles allowed us to
consider battles where the hydra can make an arbitrary number of copies at any time, which make easier our proof by absurd.
Unfortunately, the examples  most shown in the litterature
(see for instance \cite{KP82, bauer2008, BauerHydra}) 
assume that the hydra grows $n$ copies at step $n$ of the game, which is incompatible with our proof:

\vspace{6pt}

We prove now that Theorem~\ref{thm3} still holds with this ``usual'' kind of battles, by borrowing new combinatorial results from~\cite{KS81}.

 Without loss of generality, we could consider the following restrictions:

 \begin{itemize}
   \item The game starts at an initial step $i\in\mathbb{N}$ (not necessariliy $0$).
   \item  The initial hydra is the graphical representation of an ordinal  $\alpha<\epsilon_0$  in Cantor normal form (See for instance Fig~\ref{fig:iota-example}). 
 \item Hercules always chops off the rightmost head of the hydra.
 \end{itemize}

 It mathematical terms, if at step $n$ the hydra is associated with the ordinal $\alpha$, at step $n+1$ it is associated with
 $\canonseq{\alpha}{n+1}$, the $(n+1)$th element of the canonical sequence of $\alpha$~\cite{KS81}.

@@

  \item Presentation of the main steps of proofs, possibly interleaved with text (Figures \ref{fig:Ex42E0} and \ref{fig:Ex42-schutte}).

@@

Another direction would be to write in \textit{Coq} a formal proof of the original statement of Theorem~\ref{kp:thm2}. For this purpose, we plan to import  the formalization of Peano Arithmetic by O'Connor~\cite{Goedel}.

### 2021-09-22 doc/hydras.tex
comparison with the earlier (2018) paper

\author{Pierre Castéran\\ LaBRI, Univ. Bordeaux, CNRS UMR 5800, \thanks{With contributions by Yves Bertot, \'Evelyne Contejean,  Jérémy Damour, Florian Hatat, Pascal Manoury, Karl Palmskog, Clément Pit-Claudel, and Théo Zimmermann. The formalization of primitive recursive functions was originally authored by 
Russel O'Connor\cite{OConnor05}.}}
\date{\today}
%\date{August 2021}

>>>>

\author{Pierre Castéran\\ Univ. Bordeaux, CNRS, Bordeaux INP, LaBRI, UMR 5800, F-33400 Talence, France \thanks{With contributions by Yves Bertot, \'Evelyne Contejean,  Jérémy Damour, Florian Hatat, Pascal Manoury, Karl Palmskog, Clément Pit-Claudel, and Théo Zimmermann. The formalization of primitive recursive functions was originally authored by 
Russel O'Connor\cite{OConnor05}.}}
\date{\today}
%\date{August 2021}
